From Causal Discovery to Intervention Simulation: Modeling Precariousness and Depression in the HELIUS Study

### 2025-08-12 draft/draft_v2.qmd
update_folder

affiliations:
  - id: 1
    name: \textit{Computational Science Lab, Informatics Institute, University of Amsterdam, PO Box 94323, Amsterdam, 1090GH, the Netherlands}
  - id: 2
    name: \textit{Department of Public and Occupational Health, Amsterdam Public Health Research Institute, Amsterdam UMC, University of Amsterdam, Amsterdam, the Netherland}
  - id: 3
    name: \textit{Institute for Advanced Study, University of Amsterdam, Oude Turfmarkt 147, Amsterdam, 1012GC, the Netherland}   
format: 
  # docx:
  #   number-sections: true
  #   appendix: true
  pdf:
    pdf-engine: pdflatex
    fontfamily: palatino
    toc: true
    number-sections: true
    appendix: true
    colorlinks: true
    keep-tex: true
    template-partials:
      - title.tex
    include-in-header: header.tex
    crossref:
      appendix-title: "App."
      appendix-delim: ":"
date: today
execute:
  echo: false
  warning: false
bibliography: references_draft.bib
nocite: |
  @*
csl: apa.csl
abstract: |
  \noindent This study investigates how the internal structure of depression–precariousness dynamics shapes responsiveness to external support. Using cross-sectional data from the HELIUS study, we first apply cycle-capable causal discovery algorithms to estimate directional dependencies between financial stress, domain-specific precariousness, and depressive symptoms. Our multi-resolution analysis—spanning both aggregate and symptom-level constructs—highlights financial stress as a plausible upstream driver of depression, with symptoms such as sleep disturbance and depressed mood emerging as initiators of broader social vulnerability.
  To translate these structural insights into dynamic predictions, we construct a family of analytically calibrated linear stochastic models. Simulations reveal that variations in internal structure can substantially alter system responses to financial stress reduction. Some models show rapid but transient improvements in depression, while others adjust more slowly yet sustain gains over time. Precariousness tends to respond more immediately across all configurations, though its trajectory also shifts subtly with structural variation.
  Together, these findings demonstrate that intervention outcomes depend not only on the magnitude of external input but also on latent properties of the system. More broadly, this study illustrates how combining causal discovery with dynamical modeling supports principled, simulation-based exploration of intervention scenarios—grounded in data, yet able to probe counterfactual structure. This approach offers a scalable framework for diagnosing hidden system constraints and informing targeted public health strategies in complex psychosocial domains.

@@

```{r}
#| label: fig-intervention
#| fig-cap: >
#|   System response to external intervention across variations in internal structure.
#|   ***Panel A**: Mean levels of depression ($D$) and precariousness ($P$) at time steps 100, 300, and 500, across varying values of $\alpha_{DP}$ and intervention strengths. The x-axis represents the strength of external support, operationalized as the proportional reduction of financial stress ($S$) toward its minimum observed value. Each line corresponds to a different model defined by a specific $\alpha_{DP}$ value, color-coded from low (purple) to high (yellow). Models with lower $\alpha_{DP}$ exhibit faster early reductions in $D$, while higher $\alpha_{DP}$ leads to more immediate declines in $P$.
#|   
#|   **Panel B**: Full time trajectories for two models—one with no coupling ($\alpha_{DP} = 0$) and one with strong coupling ($\alpha_{DP} = 0.65$)—illustrating system response and recovery following intervention onset (step 300) and withdrawal (step 1500).
#|   Depression responds more gradually and persistently in the high $\alpha_{DP}$ model, whereas precariousness adjusts more quickly.*

#| fig-align: center
#| out-width: 100%

@@

@@

This study combined constraint-based causal discovery and dynamical modeling to investigate how depression and socioeconomic precariousness interact and how their internal structure modulates responsiveness to intervention.
Consistent with prior work linking financial strain to psychological distress [@lund2010poverty; @ezzy1993unemployment], our causal analysis identified financial stress as a key upstream driver of depression—emerging across both aggregated and disaggregated models. This provides empirical support for longstanding hypotheses that economic stressors can act as modifiable entry points into broader cycles of psychosocial risk.

At the symptom level, our analysis aligns with established theories of psychopathology but offers a more fine-grained map of potential causal relationships, generating concrete, testable hypotheses. It also points to possible limitations of prevailing models, as some observed symptom–symptom links may be shaped by shared confounders rather than direct causal influence. This suggests that, in some contexts, broader mechanisms or clusters of symptoms may provide more effective units of analysis and intervention than individual symptoms alone. 
Within this framework, sleep disturbance emerged as a potentially initiating symptom, with directed connections to both symptom-level and precariousness variables—mirroring longitudinal findings that identify sleep disruption as an early warning signal for depressive episodes and social withdrawal [@zawadzki2013rumination; @alvaro2013systematic].
Depressed mood, another core symptom, occupied a central position in the network. Not only was it influenced by upstream symptoms like anhedonia and guilt, but it also showed directional influence on social precariousness—suggesting that internal psychological states may actively shape external vulnerability. The CCI algorithm further identified a potential feedback loop between depressed mood and overall precariousness, resonating with social drift theories in depression, which posit that declining mental health can lead to social exclusion, job instability, and housing insecurity, reinforcing depressive cycles [@compton2015social; @boschloo2015network].

In contrast, social precariousness itself appeared more often as an effect than a cause, receiving arrows from symptoms like depressed mood and guilt but rarely sending them. This asymmetry supports prior observations that indicators of social vulnerability often reflect downstream consequences of poor mental health, though the relationships are likely bidirectional over time [@lund2010poverty]. 
These patterns, while derived from Amsterdam-specific data, align with mental health research in other high-income city settings, where structural inequalities such as poverty, unemployment, and housing insecurity are persistent drivers of poor psychological outcomes [@fryers2003social; @weich1998poverty].
Although the specific dynamics may vary by context, the parallels with broader literature suggest that similar mechanisms could operate in comparable settings. Nonetheless, replication in other populations will be important for assessing robustness and external validity.

Taken together, these findings yield actionable insights for mental health intervention. Symptoms like sleep disturbance, which demonstrate early and broad influence across both symptom and precariousness networks, may serve as effective targets for prevention—potentially halting the progression into more entrenched depressive states or worsening precariousness. Similarly, intervening on central nodes like depressed mood may offer strategic leverage for breaking mutually reinforcing cycles of mental decline and social isolation.
Importantly, interventions targeting such nodes may have ripple effects across domains. For instance, evidence suggests that treating core depressive symptoms can improve social functioning, employment engagement, and housing stability—particularly when combined with social or financial supports [@mcdaid2019economic; @lund2018social]. These dual benefits underscore the value of integrated, cross-domain strategies that simultaneously address internal symptoms and external stressors.

From a systems perspective, our results also highlight why single-point interventions may fail to produce lasting change. The broader causal structure reveals multiple interlocking feedback loops spanning both psychological symptoms and structural risk factors, which may re-amplify vulnerability even after temporary relief. Supporting this, recent simulation work by @park2025role shows that feedback loops, even when modest in strength, can substantially diminish the long-term impact of isolated interventions unless multiple reinforcing links are simultaneously disrupted.
Our own simulations support this view: while all models converged to similar long-term outcomes under sustained support, differences in internal structure—specifically, the strength of the link from precariousness to depression—produced systematic variation in how quickly and through which pathways systems responded to intervention. Models with stronger internal coupling exhibited slower recovery dynamics and greater persistence of depression following stress reduction, suggesting that internal system architecture, not just external input, shapes the effectiveness of support.
This supports growing calls for system-aware intervention design [@borsboom2017network]. Rather than assuming uniform responses, more durable outcomes may require (1) targeting early-acting symptoms, (2) sustaining support where recovery is slow, and (3) combining structural and psychological interventions. Tailoring strategies to match the system’s architecture can improve effectiveness, particularly in populations with layered vulnerabilities.
In this light, understanding causal architecture is not merely theoretical—it is foundational to designing more effective and equitable interventions. By identifying key modifiable drivers like financial stress and influential symptoms like sleep disturbance, and by modeling how system dynamics unfold under intervention, our study offers a flexible framework for improving mental health outcomes in complex social environments.

>>>>

affiliations:
  - id: 1
    name: \textit{Computational Science Lab, Informatics Institute, University of Amsterdam, PO Box 94323, Amsterdam, 1090GH, the Netherlands}
  - id: 2
    name: \textit{Department of Public and Occupational Health, Amsterdam Public Health Research Institute, Amsterdam UMC, University of Amsterdam, Amsterdam, the Netherland}
  - id: 3
    name: \textit{Institute for Advanced Study, University of Amsterdam, Oude Turfmarkt 147, Amsterdam, 1012GC, the Netherland}   
format: 
  # docx:
  #   number-sections: true
  #   appendix: true
  pdf:
    pdf-engine: pdflatex
    fontfamily: palatino
    toc: true
    number-sections: true
    appendix: true
    colorlinks: true
    keep-tex: true
    geometry: 
      #- showframe
      - top=40mm
      - left=45mm
      - right=45mm
      - bottom=40mm
    template-partials:
      - title.tex
    include-in-header: header.tex
    crossref:
      appendix-title: "App."
      appendix-delim: ":"
date: today
execute:
  echo: false
  warning: false
bibliography: references_draft.bib
nocite: |
  @*
csl: apa.csl
abstract: |
  \noindent This study investigates how the internal structure of depression–precariousness dynamics shapes responsiveness to external support. Using cross-sectional data from the HELIUS study, we first apply cycle-capable causal discovery algorithms to estimate directional dependencies between financial stress, domain-specific precariousness, and depressive symptoms. Our multi-resolution analysis—spanning both aggregate and symptom-level constructs—highlights financial stress as a plausible upstream driver of depression, with symptoms such as sleep disturbance and depressed mood emerging as initiators of broader social vulnerability.
  To translate these structural insights into dynamic predictions, we construct a family of analytically calibrated linear stochastic models. Simulations reveal that variations in internal structure can substantially alter system responses to financial stress reduction. Some models show rapid but transient improvements in depression, while others adjust more slowly yet sustain gains over time. Precariousness tends to respond more immediately across all configurations, though its trajectory also shifts subtly with structural variation.
  Together, these findings demonstrate that intervention outcomes depend not only on the magnitude of external input but also on latent properties of the system. More broadly, this study illustrates how combining causal discovery with dynamical modeling supports principled, simulation-based exploration of intervention scenarios—grounded in data, yet able to probe counterfactual structure. This approach offers a scalable framework for diagnosing hidden system constraints and informing targeted public health strategies in complex psychosocial domains.

@@

\newgeometry{top=20mm, bottom=20mm, left=40mm, right=40mm}

```{r}
#| label: fig-intervention
#| fig-align: center
#| out-width: 80%
#| fig-cap: >
#|   System response to external intervention across variations in internal structure.
#|   ***Panel A**: Mean levels of depression ($D$) and precariousness ($P$) at time steps 100, 300, and 500, across varying values of $\alpha_{DP}$ and intervention strengths. The x-axis represents the strength of external support, operationalized as the proportional reduction of financial stress ($S$) toward its minimum observed value. Each line corresponds to a different model defined by a specific $\alpha_{DP}$ value, color-coded from low (purple) to high (yellow). Models with lower $\alpha_{DP}$ exhibit faster early reductions in $D$, while higher $\alpha_{DP}$ leads to more immediate declines in $P$.
#|   
#|   **Panel B**: Full time trajectories for two models—one with no coupling ($\alpha_{DP} = 0$) and one with strong coupling ($\alpha_{DP} = 0.65$)—illustrating system response and recovery following intervention onset (step 300) and withdrawal (step 1500).
#|   Depression responds more gradually and persistently in the high $\alpha_{DP}$ model, whereas precariousness adjusts more quickly.*

@@

\restoregeometry

@@

This study combined constraint-based causal discovery and dynamical modeling to investigate how depression and socioeconomic precariousness interact and how their internal structure modulates responsiveness to intervention.
Consistent with prior work linking financial strain to psychological distress [@lund2010poverty; @ezzy1993unemployment], our analysis identified financial stress as a key upstream driver of depression in both aggregated and disaggregated models. This supports longstanding hypotheses that economic stressors can act as modifiable entry points into broader cycles of psychosocial risk.

At the symptom level, our analysis aligns with established theories of psychopathology but provides a more fine-grained map of potential causal relationships, generating concrete, testable hypotheses. It also points to possible limitations of prevailing models, as some observed symptom–symptom links may be shaped by shared confounders rather than direct causal influence. This suggests that, in some contexts, broader mechanisms or clusters of symptoms may provide more effective units of analysis and intervention than individual symptoms alone. 
Within this framework, sleep disturbance emerged as a potentially initiating symptom, with directed connections to both symptom-level and precariousness variables—mirroring longitudinal findings that identify sleep disruption as an early warning signal for depressive episodes and social withdrawal [@zawadzki2013rumination; @alvaro2013systematic].
Depressed mood, another core symptom, occupied a central position in the network. Not only was it influenced by upstream symptoms like anhedonia and guilt, but it also showed directional influence on social precariousness—suggesting that internal psychological states may actively shape external vulnerability. The CCI algorithm further identified a potential feedback loop between depressed mood and overall precariousness, resonating with social drift theories in depression, which posit that declining mental health can lead to social exclusion, job instability, and housing insecurity, reinforcing depressive cycles [@compton2015social; @boschloo2015network].

In contrast, social precariousness itself appeared more often as an effect than a cause, receiving arrows from symptoms like depressed mood and guilt but rarely sending them. This asymmetry supports prior observations that indicators of social vulnerability often reflect downstream consequences of poor mental health, though the relationships are likely bidirectional over time [@lund2010poverty]. 
Although our findings are based on Amsterdam-specific data, they align with research in other high-income urban settings, where structural inequalities such as poverty, unemployment, and housing insecurity consistently predict poor psychological outcomes [@fryers2003social; @weich1998poverty]. While local contexts may alter specific pathways, these parallels suggest that similar mechanisms could operate elsewhere, though replication in other populations is needed to assess robustness.

Taken together, these findings yield actionable insights for mental health intervention. Symptoms like sleep disturbance, which demonstrate early and broad influence across both symptom and precariousness networks, may serve as effective targets for prevention—potentially halting the progression into more entrenched depressive states or worsening precariousness. Similarly, targeting symptoms such as depressed mood, which appear embedded in potential feedback loops linking psychological and social domains, may offer strategic leverage for breaking mutually reinforcing cycles of mental decline and social isolation.
Importantly, interventions targeting such nodes may have ripple effects across domains. For instance, evidence suggests that treating core depressive symptoms can improve social functioning, employment engagement, and housing stability—particularly when combined with social or financial supports [@mcdaid2019economic; @lund2018social]. These dual benefits underscore the value of integrated, cross-domain strategies that simultaneously address internal symptoms and external stressors.

Our results also highlight why single-point interventions may fail to produce lasting change. The broader causal structure reveals multiple interlocking feedback loops spanning both psychological symptoms and structural risk factors, which may re-amplify vulnerability even after temporary relief. Supporting this, recent simulation work by @park2025role shows that feedback loops, even when modest in strength, can substantially diminish the long-term impact of isolated interventions unless multiple reinforcing links are simultaneously disrupted.
Our own simulations support this view: while all models converged to similar long-term outcomes under sustained support, differences in internal structure—specifically, the strength of the link from precariousness to depression—produced systematic variation in how quickly and through which pathways systems responded to intervention. Stronger coupling slowed recovery and prolonged depression, showing that internal architecture can be as important as external input.
This supports growing calls for systems-oriented mental health perspectives [@borsboom2017network]. Rather than assuming uniform responses, more durable outcomes may require (1) targeting early-acting symptoms, (2) sustaining support where recovery is slow, and (3) combining structural and psychological interventions. Tailoring strategies to match the system’s architecture can improve effectiveness, particularly in populations with layered vulnerabilities.
In this light, understanding causal architecture is not merely theoretical—it is foundational to designing more effective and equitable interventions. 
<!-- By identifying key modifiable drivers like financial stress and influential symptoms like sleep disturbance, and by modeling how system dynamics unfold under intervention, our study offers a flexible framework for improving mental health outcomes in complex social environments. -->

### 2025-08-12 draft/draft_v2.tex
update_folder

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage[]{palatino}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\usepackage{amssymb}
\usepackage{booktabs} % Add to your preamble for cleaner table lines
\usepackage{makecell} % Add to your preamble for multi-line cells
\usepackage{tikz}
\usepackage{amsmath}
\usetikzlibrary{arrows.meta, shapes.geometric}

@@

@@

\includegraphics{img/intervention_linear.pdf}

@@

This study combined constraint-based causal discovery and dynamical
modeling to investigate how depression and socioeconomic precariousness
interact and how their internal structure modulates responsiveness to
intervention. Consistent with prior work linking financial strain to
psychological distress (Ezzy, 1993; Lund et al., 2010), our causal
analysis identified financial stress as a key upstream driver of
depression---emerging across both aggregated and disaggregated models.
This provides empirical support for longstanding hypotheses that
economic stressors can act as modifiable entry points into broader
cycles of psychosocial risk.

At the symptom level, our analysis aligns with established theories of
psychopathology but offers a more fine-grained map of potential causal
relationships, generating concrete, testable hypotheses. It also points
to possible limitations of prevailing models, as some observed
symptom--symptom links may be shaped by shared confounders rather than
direct causal influence. This suggests that, in some contexts, broader
mechanisms or clusters of symptoms may provide more effective units of
analysis and intervention than individual symptoms alone. Within this
framework, sleep disturbance emerged as a potentially initiating
symptom, with directed connections to both symptom-level and
precariousness variables---mirroring longitudinal findings that identify
sleep disruption as an early warning signal for depressive episodes and
social withdrawal (Alvaro et al., 2013; Zawadzki et al., 2013).
Depressed mood, another core symptom, occupied a central position in the
network. Not only was it influenced by upstream symptoms like anhedonia
and guilt, but it also showed directional influence on social
precariousness---suggesting that internal psychological states may
actively shape external vulnerability. The CCI algorithm further
identified a potential feedback loop between depressed mood and overall
precariousness, resonating with social drift theories in depression,
which posit that declining mental health can lead to social exclusion,
job instability, and housing insecurity, reinforcing depressive cycles
(Boschloo et al., 2015; Compton \& Shim, 2015).

In contrast, social precariousness itself appeared more often as an
effect than a cause, receiving arrows from symptoms like depressed mood
and guilt but rarely sending them. This asymmetry supports prior
observations that indicators of social vulnerability often reflect
downstream consequences of poor mental health, though the relationships
are likely bidirectional over time (Lund et al., 2010). These patterns,
while derived from Amsterdam-specific data, align with mental health
research in other high-income city settings, where structural
inequalities such as poverty, unemployment, and housing insecurity are
persistent drivers of poor psychological outcomes (Fryers et al., 2003;
Weich \& Lewis, 1998). Although the specific dynamics may vary by
context, the parallels with broader literature suggest that similar
mechanisms could operate in comparable settings. Nonetheless,
replication in other populations will be important for assessing
robustness and external validity.

Taken together, these findings yield actionable insights for mental
health intervention. Symptoms like sleep disturbance, which demonstrate
early and broad influence across both symptom and precariousness
networks, may serve as effective targets for prevention---potentially
halting the progression into more entrenched depressive states or
worsening precariousness. Similarly, intervening on central nodes like
depressed mood may offer strategic leverage for breaking mutually
reinforcing cycles of mental decline and social isolation. Importantly,
interventions targeting such nodes may have ripple effects across
domains. For instance, evidence suggests that treating core depressive
symptoms can improve social functioning, employment engagement, and
housing stability---particularly when combined with social or financial
supports (Lund et al., 2018; McDaid et al., 2019). These dual benefits
underscore the value of integrated, cross-domain strategies that
simultaneously address internal symptoms and external stressors.

From a systems perspective, our results also highlight why single-point
interventions may fail to produce lasting change. The broader causal
structure reveals multiple interlocking feedback loops spanning both
psychological symptoms and structural risk factors, which may re-amplify
vulnerability even after temporary relief. Supporting this, recent
simulation work by Park, Li, et al. (2025) shows that feedback loops,
even when modest in strength, can substantially diminish the long-term
impact of isolated interventions unless multiple reinforcing links are
simultaneously disrupted. Our own simulations support this view: while
all models converged to similar long-term outcomes under sustained
support, differences in internal structure---specifically, the strength
of the link from precariousness to depression---produced systematic
variation in how quickly and through which pathways systems responded to
intervention. Models with stronger internal coupling exhibited slower
recovery dynamics and greater persistence of depression following stress
reduction, suggesting that internal system architecture, not just
external input, shapes the effectiveness of support. This supports
growing calls for system-aware intervention design (Borsboom, 2017).
Rather than assuming uniform responses, more durable outcomes may
require (1) targeting early-acting symptoms, (2) sustaining support
where recovery is slow, and (3) combining structural and psychological
interventions. Tailoring strategies to match the system's architecture
can improve effectiveness, particularly in populations with layered
vulnerabilities. In this light, understanding causal architecture is not
merely theoretical---it is foundational to designing more effective and
equitable interventions. By identifying key modifiable drivers like
financial stress and influential symptoms like sleep disturbance, and by
modeling how system dynamics unfold under intervention, our study offers
a flexible framework for improving mental health outcomes in complex
social environments.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage[]{palatino}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=40mm,left=45mm,right=45mm,bottom=40mm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\usepackage{amssymb}
\usepackage{booktabs} % Add to your preamble for cleaner table lines
\usepackage{makecell} % Add to your preamble for multi-line cells
\usepackage{tikz}
\usepackage{amsmath}
\usepackage{geometry}

\usetikzlibrary{arrows.meta, shapes.geometric}

@@

\newgeometry{top=20mm, bottom=20mm, left=40mm, right=40mm}

@@

\includegraphics[width=0.8\textwidth,height=\textheight]{img/intervention_linear.pdf}

@@

This study combined constraint-based causal discovery and dynamical
modeling to investigate how depression and socioeconomic precariousness
interact and how their internal structure modulates responsiveness to
intervention. Consistent with prior work linking financial strain to
psychological distress (Ezzy, 1993; Lund et al., 2010), our analysis
identified financial stress as a key upstream driver of depression in
both aggregated and disaggregated models. This supports longstanding
hypotheses that economic stressors can act as modifiable entry points
into broader cycles of psychosocial risk.

At the symptom level, our analysis aligns with established theories of
psychopathology but provides a more fine-grained map of potential causal
relationships, generating concrete, testable hypotheses. It also points
to possible limitations of prevailing models, as some observed
symptom--symptom links may be shaped by shared confounders rather than
direct causal influence. This suggests that, in some contexts, broader
mechanisms or clusters of symptoms may provide more effective units of
analysis and intervention than individual symptoms alone. Within this
framework, sleep disturbance emerged as a potentially initiating
symptom, with directed connections to both symptom-level and
precariousness variables---mirroring longitudinal findings that identify
sleep disruption as an early warning signal for depressive episodes and
social withdrawal (Alvaro et al., 2013; Zawadzki et al., 2013).
Depressed mood, another core symptom, occupied a central position in the
network. Not only was it influenced by upstream symptoms like anhedonia
and guilt, but it also showed directional influence on social
precariousness---suggesting that internal psychological states may
actively shape external vulnerability. The CCI algorithm further
identified a potential feedback loop between depressed mood and overall
precariousness, resonating with social drift theories in depression,
which posit that declining mental health can lead to social exclusion,
job instability, and housing insecurity, reinforcing depressive cycles
(Boschloo et al., 2015; Compton \& Shim, 2015).

In contrast, social precariousness itself appeared more often as an
effect than a cause, receiving arrows from symptoms like depressed mood
and guilt but rarely sending them. This asymmetry supports prior
observations that indicators of social vulnerability often reflect
downstream consequences of poor mental health, though the relationships
are likely bidirectional over time (Lund et al., 2010). Although our
findings are based on Amsterdam-specific data, they align with research
in other high-income urban settings, where structural inequalities such
as poverty, unemployment, and housing insecurity consistently predict
poor psychological outcomes (Fryers et al., 2003; Weich \& Lewis, 1998).
While local contexts may alter specific pathways, these parallels
suggest that similar mechanisms could operate elsewhere, though
replication in other populations is needed to assess robustness.

Taken together, these findings yield actionable insights for mental
health intervention. Symptoms like sleep disturbance, which demonstrate
early and broad influence across both symptom and precariousness
networks, may serve as effective targets for prevention---potentially
halting the progression into more entrenched depressive states or
worsening precariousness. Similarly, targeting symptoms such as
depressed mood, which appear embedded in potential feedback loops
linking psychological and social domains, may offer strategic leverage
for breaking mutually reinforcing cycles of mental decline and social
isolation. Importantly, interventions targeting such nodes may have
ripple effects across domains. For instance, evidence suggests that
treating core depressive symptoms can improve social functioning,
employment engagement, and housing stability---particularly when
combined with social or financial supports (Lund et al., 2018; McDaid et
al., 2019). These dual benefits underscore the value of integrated,
cross-domain strategies that simultaneously address internal symptoms
and external stressors.

Our results also highlight why single-point interventions may fail to
produce lasting change. The broader causal structure reveals multiple
interlocking feedback loops spanning both psychological symptoms and
structural risk factors, which may re-amplify vulnerability even after
temporary relief. Supporting this, recent simulation work by Park, Li,
et al. (2025) shows that feedback loops, even when modest in strength,
can substantially diminish the long-term impact of isolated
interventions unless multiple reinforcing links are simultaneously
disrupted. Our own simulations support this view: while all models
converged to similar long-term outcomes under sustained support,
differences in internal structure---specifically, the strength of the
link from precariousness to depression---produced systematic variation
in how quickly and through which pathways systems responded to
intervention. Stronger coupling slowed recovery and prolonged
depression, showing that internal architecture can be as important as
external input. This supports growing calls for systems-oriented mental
health perspectives (Borsboom, 2017). Rather than assuming uniform
responses, more durable outcomes may require (1) targeting early-acting
symptoms, (2) sustaining support where recovery is slow, and (3)
combining structural and psychological interventions. Tailoring
strategies to match the system's architecture can improve effectiveness,
particularly in populations with layered vulnerabilities. In this light,
understanding causal architecture is not merely theoretical---it is
foundational to designing more effective and equitable interventions.
